---
title: "IAS Data Exploration"
subtitle: "Initial Analysis for Virtual Paper Project"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# Overview

This notebook explores the IAS officer data to determine project viability.

**Key questions:**
1. Do we have variation in educational backgrounds?
2. How many District Collector positions exist?
3. Can we merge the three datasets cleanly?
4. Is there enough data for analysis?
```{r load-packages}
library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)

# Set theme for plots
theme_set(theme_minimal())
```

---

# 1. Load Data
```{r load-data}
# Load the three CSV files
profile <- read_csv("../data/raw/ias-profile.csv")
education <- read_csv("../data/raw/ias-education.csv")
experience <- read_csv("../data/raw/ias-experience.csv")

# Quick look at dimensions
cat("Profile data:", nrow(profile), "officers\n")
cat("Education data:", nrow(education), "rows (multiple degrees per officer)\n")
cat("Experience data:", nrow(experience), "rows (multiple positions per officer)\n")
```

---

# 2. Profile Data Exploration
```{r profile-overview}
# What's in the profile data?
glimpse(profile)

# Key variables
cat("\nKey variables in profile:\n")
cat("- ID (unique identifier)\n")
cat("- Name\n")
cat("- Cadre (state assignment)\n")
cat("- Allotment_Year (year joined IAS)\n")
cat("- Gender\n")
cat("- Retired (yes/no)\n")
```

## Distribution of officers over time
```{r profile-time}
# Officers by allotment year
profile %>%
  filter(!is.na(Allotment_Year), Allotment_Year >= 1950) %>%
  count(Allotment_Year) %>%
  ggplot(aes(x = Allotment_Year, y = n)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue") +
  labs(
    title = "IAS Officers by Allotment Year",
    subtitle = "1950-2020",
    x = "Allotment Year",
    y = "Number of Officers"
  )
```

## Cadre distribution
```{r profile-cadre}
# Top 10 cadres (states)
profile %>%
  count(Cadre, sort = TRUE) %>%
  head(10) %>%
  kable(col.names = c("State Cadre", "Number of Officers")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Key finding:** Officers are distributed across states. We can use within-state, across-district variation.

---

# 3. Education Data Exploration
```{r education-overview}
glimpse(education)

cat("\nKey variables in education:\n")
cat("- ID (links to profile)\n")
cat("- Qualification (degree type)\n")
cat("- Subject (specific field)\n")
cat("- Category_of_Subject (pre-coded categories)\n")
cat("- Division (First Class, Second Class, etc.)\n")
```

## **CRITICAL QUESTION: Educational Background Variation**
```{r education-variation}
# Distribution by category
edu_dist <- education %>%
  count(Category_of_Subject, sort = TRUE) %>%
  mutate(
    pct = n / sum(n) * 100,
    pct_label = paste0(round(pct, 1), "%")
  )

# Show table
edu_dist %>%
  kable(
    col.names = c("Subject Category", "Count", "Percentage", "Label"),
    digits = 1
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Plot
ggplot(edu_dist, aes(x = reorder(Category_of_Subject, n), y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = pct_label), hjust = -0.1) +
  coord_flip() +
  labs(
    title = "Distribution of Educational Backgrounds",
    subtitle = "All IAS Officers, 1951-2020",
    x = NULL,
    y = "Number of Officers"
  )
```

### Focus on key categories
```{r education-key}
# Engineers vs Social Scientists
key_cats <- education %>%
  filter(Category_of_Subject %in% c(
    "Engineering, Technology and Mathematics",
    "Social Sciences and Liberal Arts"
  )) %>%
  count(Category_of_Subject) %>%
  mutate(pct = n / sum(n) * 100)

key_cats %>%
  kable(
    col.names = c("Category", "Count", "Percentage"),
    digits = 1
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Assessment:**
```{r education-assessment}
min_count <- min(key_cats$n)
if (min_count < 500) {
  cat("⚠️  WARNING: Small sample size in one category!\n")
  cat("   Minimum count:", min_count, "\n")
  cat("   Project may have limited statistical power.\n")
} else if (min_count < 1000) {
  cat("⚠️  CAUTION: Moderate sample size.\n")
  cat("   Minimum count:", min_count, "\n")
  cat("   Project is viable but consider power calculations.\n")
} else {
  cat("✓ GOOD: Sufficient variation for robust analysis.\n")
  cat("   Minimum count:", min_count, "\n")
}
```

### Trends over time
```{r education-trends}
# Join with profile to get allotment year
edu_time <- education %>%
  left_join(
    profile %>% select(ID, Allotment_Year),
    by = "ID"
  ) %>%
  filter(
    !is.na(Allotment_Year),
    Allotment_Year >= 1990,
    Allotment_Year <= 2020,
    Category_of_Subject %in% c(
      "Engineering, Technology and Mathematics",
      "Social Sciences and Liberal Arts",
      "Natural Sciences",
      "Governance, Law and Management"
    )
  ) %>%
  count(Allotment_Year, Category_of_Subject)

ggplot(edu_time, aes(x = Allotment_Year, y = n, color = Category_of_Subject)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Educational Backgrounds Over Time",
    subtitle = "IAS Officers, 1990-2020",
    x = "Allotment Year",
    y = "Number of Officers",
    color = "Subject Category"
  ) +
  scale_color_brewer(palette = "Set1")
```

---

# 4. Experience Data Exploration
```{r experience-overview}
glimpse(experience)

cat("\nKey variables in experience:\n")
cat("- ID (links to profile)\n")
cat("- Designation (job title)\n")
cat("- Office (location/organization)\n")
cat("- Start_Date, End_Date\n")
cat("- Field_of_Experience (sector)\n")
```

## District Collector positions
```{r dc-positions}
# Identify District Collector positions
dc_data <- experience %>%
  filter(
    str_detect(Designation, regex("District Collector|District Magistrate", ignore_case = TRUE))
  ) %>%
  mutate(
    Start_Date = ymd(Start_Date),
    End_Date = ymd(End_Date),
    year = year(Start_Date),
    tenure_days = as.numeric(End_Date - Start_Date),
    tenure_months = tenure_days / 30
  )

cat("Total District Collector position records:", nrow(dc_data), "\n")

# Remove invalid dates
dc_data_clean <- dc_data %>%
  filter(
    !is.na(Start_Date),
    !is.na(End_Date),
    tenure_days > 0,
    tenure_days < 3650,  # Less than 10 years
    year >= 1990,
    year <= 2020
  )

cat("DC positions with valid dates (1990-2020):", nrow(dc_data_clean), "\n")
```

### Tenure distribution
```{r dc-tenure}
# Summary stats
tenure_summary <- summary(dc_data_clean$tenure_months)
print(tenure_summary)

# Histogram
ggplot(dc_data_clean, aes(x = tenure_months)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(
    xintercept = median(dc_data_clean$tenure_months, na.rm = TRUE),
    linetype = "dashed",
    color = "red",
    size = 1
  ) +
  annotate(
    "text",
    x = median(dc_data_clean$tenure_months, na.rm = TRUE) + 5,
    y = Inf,
    label = paste("Median:", round(median(dc_data_clean$tenure_months, na.rm = TRUE), 1), "months"),
    vjust = 2,
    color = "red"
  ) +
  labs(
    title = "Distribution of District Collector Tenure",
    subtitle = "1990-2020",
    x = "Tenure (months)",
    y = "Count"
  )
```

### DC positions over time
```{r dc-time}
dc_time <- dc_data_clean %>%
  count(year)

ggplot(dc_time, aes(x = year, y = n)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Number of District Collector Positions Per Year",
    subtitle = "1990-2020",
    x = "Year",
    y = "Number of DC Positions"
  )

# Focus on 2005-2015 (NREGA era)
dc_nrega_era <- dc_data_clean %>%
  filter(year >= 2005, year <= 2015)

cat("\nDC positions in NREGA era (2005-2015):", nrow(dc_nrega_era), "\n")
```

### Can we identify districts?
```{r dc-districts}
# Sample of Office values
cat("Sample of Office values for District Collectors:\n\n")
dc_data_clean %>%
  select(ID, Designation, Office) %>%
  head(20) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Check for district keywords
dc_district_check <- dc_data_clean %>%
  mutate(
    has_district = str_detect(Office, regex("district|collectorate", ignore_case = TRUE)),
    # Try to extract district name
    potential_district = str_extract(Office, "^[A-Za-z\\s]+(?=,|District|Collectorate)")
  )

cat("\n\nDC positions with 'district' or 'collectorate' in Office:\n")
cat("Count:", sum(dc_district_check$has_district, na.rm = TRUE), "\n")
cat("Percentage:", round(mean(dc_district_check$has_district, na.rm = TRUE) * 100, 1), "%\n")

# Show unique districts
cat("\n\nSample of extracted district names:\n")
dc_district_check %>%
  filter(!is.na(potential_district)) %>%
  count(potential_district, sort = TRUE) %>%
  head(20) %>%
  kable(col.names = c("Potential District", "Count")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# 5. Merging the Three Datasets

## Strategy

We'll merge in steps:
1. Profile + Education → Get officer background
2. Result + Experience → Get career with background
3. Focus on District Collectors with education data
```{r merge-step1}
# Step 1: Get highest degree for each officer
# (Some officers have multiple degrees - take the highest)

edu_highest <- education %>%
  group_by(ID) %>%
  arrange(Reference_Value) %>%  # Lower reference value = higher degree
  slice(1) %>%
  ungroup() %>%
  select(
    ID,
    Qualification,
    Subject,
    Category_of_Subject,
    Division
  )

cat("Officers with education data:", n_distinct(edu_highest$ID), "\n")
```
```{r merge-step2}
# Step 2: Merge profile with education

officers <- profile %>%
  left_join(edu_highest, by = "ID")

cat("Officers in merged dataset:", nrow(officers), "\n")
cat("Officers with education info:", sum(!is.na(officers$Category_of_Subject)), "\n")
```
```{r merge-step3}
# Step 3: Merge with DC experience

dc_with_background <- dc_data_clean %>%
  left_join(
    officers %>% select(ID, Allotment_Year, Cadre, Category_of_Subject, Division),
    by = "ID"
  )

cat("\nDC positions with officer background data:\n")
cat("Total DC positions (1990-2020):", nrow(dc_with_background), "\n")
cat("DC positions with education data:", sum(!is.na(dc_with_background$Category_of_Subject)), "\n")
cat("Percentage:", round(mean(!is.na(dc_with_background$Category_of_Subject)) * 100, 1), "%\n")
```

## Focus on key backgrounds
```{r merge-final}
# Final analysis dataset: DCs with engineer or social science background

dc_analysis <- dc_with_background %>%
  filter(
    Category_of_Subject %in% c(
      "Engineering, Technology and Mathematics",
      "Social Sciences and Liberal Arts"
    )
  ) %>%
  mutate(
    background = case_when(
      Category_of_Subject == "Engineering, Technology and Mathematics" ~ "Engineer",
      Category_of_Subject == "Social Sciences and Liberal Arts" ~ "Social Scientist",
      TRUE ~ "Other"
    )
  )

cat("\n\n=== FINAL ANALYSIS DATASET ===\n")
cat("DC positions with Engineer or Social Science background:", nrow(dc_analysis), "\n")
cat("\nBreakdown by background:\n")
table(dc_analysis$background) %>%
  as.data.frame() %>%
  rename(Background = Var1, Count = Freq) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Distribution over time
```{r merge-time}
dc_analysis %>%
  count(year, background) %>%
  ggplot(aes(x = year, y = n, color = background)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "District Collector Positions by Educational Background",
    subtitle = "Engineers vs Social Scientists, 1990-2020",
    x = "Year",
    y = "Number of DC Positions",
    color = "Background"
  ) +
  scale_color_manual(values = c("Engineer" = "steelblue", "Social Scientist" = "darkgreen"))
```

### Distribution by cadre (state)
```{r merge-cadre}
dc_analysis %>%
  count(Cadre, background) %>%
  pivot_wider(names_from = background, values_from = n, values_fill = 0) %>%
  arrange(desc(Engineer + `Social Scientist`)) %>%
  head(15) %>%
  kable(caption = "Top 15 States by DC Positions") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# 6. Summary Statistics for Professor Meeting
```{r summary-table}
summary_stats <- tibble(
  Metric = c(
    "Total IAS Officers",
    "Officers with Education Data",
    "Officers - Engineers (%)",
    "Officers - Social Scientists (%)",
    "",
    "Total DC Positions",
    "DC Positions (1990-2020)",
    "DC Positions (2005-2015, NREGA era)",
    "Median DC Tenure (months)",
    "",
    "DC Positions with Education Data",
    "DC Positions - Engineers",
    "DC Positions - Social Scientists",
    "DC Positions with District Info (%)"
  ),
  Value = c(
    nrow(profile),
    n_distinct(edu_highest$ID),
    round(edu_dist$pct[edu_dist$Category_of_Subject == "Engineering, Technology and Mathematics"], 1),
    round(edu_dist$pct[edu_dist$Category_of_Subject == "Social Sciences and Liberal Arts"], 1),
    "",
    nrow(dc_data),
    nrow(dc_data_clean),
    nrow(dc_nrega_era),
    round(median(dc_data_clean$tenure_months, na.rm = TRUE), 1),
    "",
    nrow(dc_analysis),
    sum(dc_analysis$background == "Engineer"),
    sum(dc_analysis$background == "Social Scientist"),
    round(mean(dc_district_check$has_district, na.rm = TRUE) * 100, 1)
  )
)

summary_stats %>%
  kable(
    caption = "Summary Statistics for Project Viability",
    align = c("l", "r")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(c(5, 10), bold = TRUE, background = "#f0f0f0")
```

---

# 7. Project Viability Assessment
```{r viability}
# Set thresholds
MIN_SAMPLE <- 500
MIN_PERCENTAGE <- 20

engineer_count <- sum(dc_analysis$background == "Engineer")
social_sci_count <- sum(dc_analysis$background == "Social Scientist")
min_count <- min(engineer_count, social_sci_count)

district_pct <- mean(dc_district_check$has_district, na.rm = TRUE) * 100

cat("=== PROJECT VIABILITY ASSESSMENT ===\n\n")

# Sample size check
if (min_count >= MIN_SAMPLE) {
  cat("✓ SAMPLE SIZE: PASS\n")
  cat("  Smallest group has", min_count, "observations\n")
  cat("  This is sufficient for analysis.\n\n")
} else {
  cat("⚠️  SAMPLE SIZE: CONCERN\n")
  cat("  Smallest group has", min_count, "observations\n")
  cat("  This may limit statistical power.\n\n")
}

# District identification check
if (district_pct >= MIN_PERCENTAGE) {
  cat("✓ DISTRICT IDENTIFICATION: FEASIBLE\n")
  cat(" ", round(district_pct, 1), "% of DC positions have clear district info\n")
  cat("  Merge with SHRUG is viable.\n\n")
} else {
  cat("⚠️  DISTRICT IDENTIFICATION: DIFFICULT\n")
  cat("  Only", round(district_pct, 1), "% of DC positions have clear district info\n")
  cat("  Consider state-level aggregation or alternative approach.\n\n")
}

# Overall recommendation
if (min_count >= MIN_SAMPLE && district_pct >= MIN_PERCENTAGE) {
  cat("📊 OVERALL: PROJECT IS VIABLE\n")
  cat("   Recommend proceeding with DC-SHRUG merge.\n")
} else if (min_count >= MIN_SAMPLE) {
  cat("📊 OVERALL: PROJECT IS VIABLE WITH MODIFICATIONS\n")
  cat("   Recommend state-level analysis or focus on IAS careers only.\n")
} else {
  cat("⚠️  OVERALL: PROJECT NEEDS RECONSIDERATION\n")
  cat("   Consider alternative research questions or data sources.\n")
}
```

---

# 8. Next Steps

Based on this exploration:

**If viable:**
1. Clean district name extraction from Office variable
2. Download SHRUG data
3. Attempt DC-district-year merge
4. Run balance checks (are engineers in different districts?)
5. Preliminary regressions

**If not viable:**
1. Pivot to state-level analysis
2. OR focus on IAS careers (backgrounds → promotions)
3. OR consider alternative identification (event studies, political shocks)

**For professor meeting:**
1. Print summary statistics table
2. Bring figures showing distributions
3. Discuss concerns honestly
4. Get guidance on path forward

---

# Save Analysis Dataset
```{r save-data}
# Save the merged DC dataset for future analysis
saveRDS(dc_analysis, "../data/analysis/dc_with_backgrounds.rds")

cat("✓ Analysis dataset saved to: data/analysis/dc_with_backgrounds.rds\n")
cat("✓ This file contains", nrow(dc_analysis), "DC positions with background data\n")
```

---

**Analysis complete!** Review the viability assessment above and prepare for professor meeting.
